Temporal Dynamics of Normalization Reweighting

### 2022-09-30 manuscript.Rmd
Update of manuscript

---
title: 'Temporal Dynamics of Normalization Reweighting (target journal: Neuron)'
author: Daniel H. Baker, Daniela Marinova, Richard Aveyard, Lydia Hargreaves, Alice
  Renton, Ruby Castellani, Phoebe Hall, Miriam Harmens, Georgia Holroyd, Beth Nicholson,
  Emily Williams, Hannah M. Hobson & Alex R. Wade
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  word_document: default
  pdf_document:
    toc: no
  html_document: default
---

@@

processdata <- 0  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all scripts
# 1 - generate figures using the processed group data (requires XGB of storage)
# 2 - download processed individual data and analyse (requires XGB of storage)
# 3 - download all raw data and analyse (requires XGB of storage)

plotsmoothed <- 1 # either plot the raw data (0), or a smoothed spline (1)

# reasonably compact code to check which packages are installed, install the missing ones, and activate all
packagelist <- c('knitr','remotes','tictoc','R.matlab','bookdown','grImport','tiff','pals') # list of CRAN packages
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
if (!'osfr' %in% installed.packages()[,1]){remotes::install_github("centerforopenscience/osfr")}
if (!'FourierStats' %in% installed.packages()[,1]){remotes::install_github("bakerdh/FourierStats")}
packagelist <- c(packagelist,'osfr','FourierStats')
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

knitr::opts_chunk$set(echo = TRUE)
colpal <- c('#000000','#8783CF','#FE5000','#228B22','#8B8000')

@@

@@

## Abstract

For decades, neural suppression in early visual cortex has been thought to be fixed. But recent work has challenged this assumption by showing that suppression can be \emph{reweighted} based on recent history; when pairs of stimuli are repeatedly presented together, suppression between them strengthens. Here we investigate the temporal dynamics of this process using a steady-state VEP paradigm that provides a time-resolved, direct index of suppression between pairs of stimuli flickering at different frequencies (5 and 7Hz). Our initial analysis of an existing EEG dataset (N=100) indicated that suppression increases substantially during the first 4 seconds of stimulus presentation. We then collected new EEG data (N=100) replicating this finding for both monocular and dichoptic mask arrangements in a preregistered study designed to measure reweighting. A third experiment (N=20) use source localized MEG, and found that suppression increases across the visual hierarchy, and that reweighting is more rapid in dorsal compared with ventral visual areas. Because long-standing theories propose inhibition/excitation differences in autism, we also compared reweighting between individuals with high vs low autistic traits, and with and without an autism diagnosis, across our 3 data sets (total N=220). We find . . .

## Introduction

Suppressive interactions between neurons are ubiquitous in the nervous system, and normalization (or gain control) processes are considered a canonical neuronal computation (Carandini & Heeger). Yet the strength of suppression was for decades treated as fixed, largely due to the observation that adapting to one stimulus does not decrease its suppressive potency (refs). This orthodoxy was recently challenged by a series of innovative studies showing that normalization can be 'reweighted' by recent history (Westrick, Aschner, Yiltiz). Specifically, when pairs of stimuli are repeatedly presented together, they come to suppress each other more strongly. This suggests that, far from being fixed, normalization is a dynamic process that is continuously updated by the sensory environment. Here, our objectives were to determine if the timecourse of these changes can be measured non-invasively from the human brain, assess if they occur across suppressive pathways, and determine whether they differ across the population as a function of autistic traits.

Atypical sensory experience is widely reported by individuals on the autism spectrum, but the causal mechanisms remain unclear. Typical issues include hypersensitivity to intense stimuli such as loud sounds, bright lights and strong odours or flavours. Yet fundamental measures of sensitivity such as visual acuity, contrast sensitivity, and audiometric performance (Rosenhall et al. 1999) are not consistently different from neurotypical controls. Theoretical accounts of sensory differences in autism have long proposed that the balance of inhibition and excitation may be disrupted, and there are isolated results that seem consistent with this. For example 
However many other studies have failed to find group differences using tasks that are at least superficially dependent on inhibition, such as...

A comprehensive review of the literature on vision in autism (Simmons et al., 2009)

Our recent work has identified a potential autism-related difference in gain control using steady-state EEG methods. Vilidaite et al. (2018) measured contrast response functions for flickering stimuli in a large group of neurotypical adults, who were split by their autism quotient (AQ; Baron-Cohen et al., 2001) score. The responses were comparable between high and low AQ groups at the flicker frequency of the stimulus. However at the second harmonic frequency (twice the flicker frequency), the high AQ group showed weaker responses than the low AQ group. This effect was replicated in an independent sample of 12 adults with an autism diagnosis (compared with neurotypical controls). Because second harmonic responses are caused by nonlinear interactions in the visual system (they are absent in a linear system), this potentially implicates differences in divisive suppression in autism. Furthermore, there appears to be a developmental trajectory, as autistic children tested as part of the same study (Vilidaite et al., 2018) showed weaker responses at both the first and second harmonic frequencies.

Suppression itself is not a single process. Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation, eye-of-origin and spatial position. At present there is evidence of normalization reweighting between stimuli of orthogonal orientations (Aschner, 2018), and adjacent spatial positions (Yiltiz, 2020). We also wondered if interocular suppression might be subject to reweighting, and if there are differences in the dynamics across pathways. This is possible, given that suppression within and between the eyes has different spatiotemporal tuning (Meese & Baker, 2009), and dichoptic masking can be reduced by adapting to the mask (Baker, 2007 and other studies), unlike within-eye masking. Dynamic fluctuations in interocular suppression are a feature of binocular rivalry (Wilson, 2003; Alais et al., 2006), for which differences in autism have also been reported (Robertson).

We hypothesised that normalization reweighting might differ as a function of autistic traits. The relative novelty of the reweighting framework could explain why any differences have not previously been detected, and why the literature on inhibition in autism is relatively inconclusive. In this paper we perform a time-course analysis of a previously published EEG data set, and report two novel pre-registered experiments using EEG and MEG. Our data show that suppression increases substantially during the first 6 seconds after stimulus onset, for both monocular and dichoptic masks. The timecourse of reweighting is slower in ventral brain regions, and more rapid in dorsal parts of the visual pathway. Autism ...

## Results

We began by reanalysing data from a steady-state visually evoked potential (SSVEP) experiment reported by Vilidaite et al (2018). Participants viewed arrays of flickering gratings of varying contrasts. In some conditions a single grating orientation was present flickering at 7Hz, whereas in other conditions a high contrast 'mask' was added at right angles to the target gratings, and flickering at 5Hz. The left panel of Figure \@ref(fig:Pilotdata)a shows contrast response functions with and without the mask - the presence of the mask reduces the 7Hz response to the target (blue squares are below the black circles). Similarly, the right panel of Figure \@ref(fig:Pilotdata)a shows that the 5Hz response to the mask was itself suppressed by the presence of high contrast targets (note that the data from the mask conditions were not reported by Vilidaite et al.). At both frequencies, responses were localised to the occipital pole (see insets).

We then performed a timecourse analysis, in which we analysed each 11-second trial using a sliding 1-second time window. Figure \@ref(fig:Pilotdata)c shows the response at the target frequency (7Hz) to a single stimulus of 32% contrast (black), and the response at 7Hz when the mask is added (blue). Analogous responses are shown at three other frequencies - the mask frequency (5Hz), and the second harmonics of both target and mask frequencies (14Hz, 10Hz), at which strong responses were also found (see spectra in Figure \@ref(fig:Pilotdata)b). The reduction in signal strength when the mask component is added illustrates the masking effect. Taking the ratio of the two timecourses to calculate a masking index reveals that masking increases steeply during the first two seconds of stimulus presentation, and then plateaus for several seconds (blue trace in Figure \@ref(fig:Pilotdata)d). A similar pattern is observed at 5Hz (red trace in Figure \@ref(fig:Pilotdata)d), with the increase continuing for around four seconds, as well as at the second harmonics. The black trace shows the average masking ratio across all four frequencies, which rises steeply for just over two seconds, mor gradually for a further two seconds, and then stays approximately constant.

@@

@@

@@

  pdf('temp/figures/Fig2.pdf', bg="transparent", height = 6, width = 16)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

@@

```{r Pilotdata, fig.cap="Summary of pilot analysis of data from Vilidiate et al. (2018). Panel (a) shows contrast response functions at the target frequency (7Hz, left) and the mask frequency (5Hz, right). Insets show the distribution of activity across the scalp, with points marking electrodes Oz, POz, O1 and O2. Panel (b) shows Fourier spectra for the single component stimuli and their combination (plaid). Note the strong second harmonic components at 14Hz and 10Hz. Panel (c) shows timecourses of frequency-locked responses to a single stimulus (black) and the plaid stimulus (blue), compared to baseline (grey). Panel (d) shows the timecourse of suppression at each frequency (7Hz, 5Hz, 14Hz, 10Hz) and their average (black curve). Error bars in panel (a) and shaded regions in panels (c,d) indicate ±1SE, and grey rectangles indicate the timing of stimulus presentation. The larger symbols in panel (a) indicate conditions used for subsequent analyses.", fig.align="center", echo=FALSE}

@@

Our initial reanalysis was promising, however the data were noisy despite the large sample size, because each participant contributed only 8 trials (88 seconds) to each condition. We therefore preregistered two new experiments (see https://osf.io/4qudc) to investigate these effects in greater detail. These had a similar overall design to the Vilidaite study, with some small changes intended to optimise the study (see Methods). The key differences were that we used shorter trials (because there were few changes in the latter part of the trials shown in Figure \@ref(fig:Pilotdata)), and also focussed all trials into a smaller number of conditions, such that each participant contributed 48 repetitions (288 seconds of data) to each of 4 conditions.

Figure \@ref(fig:EEGdata) summarises the results of our EEG experiment testing a further 100 neurotypical adults. Averaged EEG waveforms showed a strong oscillatory component at each of the two stimulus flicker frequencies (Figure \@ref(fig:EEGdata)a), which slightly lagged the driving signal. Signals were well-isolated in the Fourier domain (Figure \@ref(fig:EEGdata)b), and localised to occipital electrodes. The timecourse at both frequencies showed an initial onset transient, and was then relatively stable for the 6 seconds of stimulus presentation (Figure \@ref(fig:EEGdata)c,d). Responses were weaker in the two masking conditions, and the ratio of target only to target + mask conditions increased over time (Figure \@ref(fig:EEGdata)e,f) for both mask types. At 5Hz the increase in masking continued over the first 5 seconds of stimulus presentation (Figure \@ref(fig:EEGdata)e), whereas at 7Hz the increase occurred mostly during the first second after onset (Figure \@ref(fig:EEGdata)f), consistent with the pilot data (see Figure \@ref(fig:Pilotdata)).

@@

allAQ[s] <- scoreAQ(qdatasubj[1,5:54])

save(file='temp/EEGsummary.RData',list=c('allheads5','allheads7','allspec','allwaves','EEGtimecourse5Hz','EEGtimecourse7Hz','electrodes','allAQ'))
 }

if (processdata>0){
load('temp/EEGsummary.RData')

@@

>>>>

---
title: 'Temporal Dynamics of Normalization Reweighting'
author: | 
  | Daniel H. Baker, Daniela Marinova, Richard Aveyard, Lydia J. Hargreaves, Alice Renton,  
  | Ruby Castellani, Phoebe Hall, Miriam Harmens, Georgia Holroyd, Beth Nicholson, 
  | Emily L. Williams, Hannah M. Hobson & Alex R. Wade
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  word_document: default
  pdf_document:
    toc: no
  html_document: default
bibliography: references.bib
csl: elife.csl
---

@@

processdata <- 0  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data (requires XGB of storage)
# 2 - download processed individual data and analyse (requires XGB of storage)
# 3 - download all raw data and analyse (requires XGB of storage)

plotsmoothed <- 1 # either plot the raw data (0), or a smoothed spline (1)
useSPQ <- 0  # optionally do the median split by SPQ for the main EEG experiment

# reasonably compact code to check which packages are installed, install the missing ones, and activate all
packagelist <- c('knitr','remotes','tictoc','R.matlab','bookdown','grImport','tiff','pals','ez','gtools') # list of CRAN packages
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
if (!'osfr' %in% installed.packages()[,1]){remotes::install_github("centerforopenscience/osfr")}
if (!'FourierStats' %in% installed.packages()[,1]){remotes::install_github("bakerdh/FourierStats")}
packagelist <- c(packagelist,'osfr','FourierStats')
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

scoreSPQ <- function(qdata){
# score the SPQ questionnaire
# input is a row from a spreadsheet indicating responses to 92 questions
SPQanswers <- toupper(c('Strongly agree','Agree','Disagree','Strongly Disagree'))
# SPQorder <- c(4,2,1,3)
reverseitemsSPQ <- (1:ncol(qdata))*0
reverseitemsSPQ[c(3,4,5,10,13,15:18,20,22,25:29,31,34,39,40,42:44,46:48,51,54,56,57,63,66,67,72,75,77:80,86,89,92)] <- 1  # reverse scored questionnaire items

  thisSPQ <- 0
  nansweredSPQ <- 0
  for (q in 1:92){
    response <- which(SPQanswers==toupper(as.character(qdata[1,q]))) - 1
    if (length(response)>0){
      nansweredSPQ <- nansweredSPQ + 1
      if (reverseitemsSPQ[q]==1){response <- 3 - response}
      thisSPQ <- thisSPQ + response
    }
  }

return(thisSPQ)}

clustercorrect <- function(datax,datay=NULL,adjacencymatrix=NULL,testtype=1,paired=TRUE,clustformthresh=0.05,clustthresh=0.05,nresamples=10000) {
  clustout <- NULL
  sigclustercount <- 0
  clusterpout <- NULL
  clusterpoints <- list()
  nulldist <- NULL
  N <- nrow(datax)  # number of observations (e.g. participants)
  m <- ncol(datax)  # number of measurements (e.g. sensors, voxels, timepoints)

  # if complex data have been passed in and t-test selected, convert to amplitudes
  if (is.complex(datax) && testtype==1){datax <- abs(datax)}
  if (is.complex(datay) && testtype==1){datay <- abs(datay)}

  # if two data sets have been passed in, note the size of the second matrix
  if (!pracma::isempty(datay)){N2 <- nrow(datay)}

  # if it's a paired design, subtract the two data sets and delete the second one
  if (paired==TRUE && !pracma::isempty(datay)){
    datax <- datax - datay
    datay <- NULL}

  if (length(adjacencymatrix)==0){
    # if no adjacency matrix has been passed in, we assume that subsequent observations are adjacent
    adjacencymatrix <- matrix(0,nrow=m,ncol=m)
    for (n in 1:(m-1)){adjacencymatrix[n,n+1] <- 1}
    for (n in 2:m){adjacencymatrix[n,n-1] <- 1}
  }

  allp <- NULL
  allt <- NULL
  # loop through all elements (sensors/time points) and calculate the test statistic and p-value
  for (n in 1:m){

    if (pracma::isempty(datay)){
      # one-sample or paired test
      if (testtype==1){
        output <- t.test(datax[,n])  # univariate t-test
        allp[n] <- output$p.value
        allt[n] <- output$statistic}
      if (testtype==2){
        output <- tsqh.test(datax[,n])  # one-sample Hotelling's t-squared
        allp[n] <- output$pval
        allt[n] <- output$tsq}
      if (testtype==3){
        output <- tsqc.test(datax[,n])  # T-squared-circ
        allp[n] <- output$pval
        allt[n] <- output$tsqc}
      if (testtype==4){
        output <- t.test(datax[,n],alternative='greater')  # univariate t-test
        allp[n] <- output$p.value
        allt[n] <- output$statistic}
    }

    # independent samples test
    if (!pracma::isempty(datay) && is.matrix(datay)){
      if (testtype==1){
        output <- t.test(datax[,n],datay[,n]) # independent t-test
        allp[n] <- output$p.value
        allt[n] <- output$statistic}
      if (testtype==2){
        output <- tsqh.test(datax[,n],datay[,n],paired=FALSE)  # independent T-squared
        allp[n] <- output$pval
        allt[n] <- output$tsq}
      if (testtype==3){
        output <- tsqc.test(datax[,n],datay[,n],paired=FALSE)  # independent T-squared-circ
        allp[n] <- output$pval
        allt[n] <- output$tsqc}
            if (testtype==4){
        output <- t.test(datax[,n],datay[,n],alternative='greater') 
        allp[n] <- output$p.value
        allt[n] <- output$statistic}
    }
  }

  allp[which(is.na(allp))] <- 1

  # now generate a list of clusters of adjacent significant elements
  hvect <- rep(0,m)
  hvect[which(allp<clustformthresh)] <- 1
  hmat <- matrix(hvect,nrow=m,ncol=m)
  hmat2 <- hmat * t(hmat)
  clustprod <- hmat2 * adjacencymatrix

  clusterlist <- list()
  nclusts <- 0
  for (n in 1:m){
    temp = sum(clustprod[n,])
    if (temp>0){
      i <- which(clustprod[n,]>0)
      tempclust <- c(n, i)
      clustprod[n,] <- 0
      clustprod[,n] <- 0
      cnt <- 0
      while (cnt<length(tempclust)){
        cnt <- cnt + 1
        temp = sum(clustprod[tempclust[cnt],])
        if (temp>0){
          i <- which(clustprod[tempclust[cnt],]>0)
          tempclust <- c(tempclust, i)
          clustprod[tempclust[cnt],] <- 0
          clustprod[,tempclust[cnt]] <- 0
        }
      }
      nclusts <- nclusts + 1
      clusterlist[[nclusts]] <- sort(unique(tempclust))
    }
  }

  if (nclusts>0){
    sumtvals <- NULL
    for (cc in 1:nclusts){
      Cindices <- as.numeric(clusterlist[[cc]])
      sumtvals[cc] <- sum(allt[Cindices])}

    i <- which(sumtvals==max(sumtvals))  # find the largest cluster
    maxcluster <- clusterlist[[i]]  # store the largest cluster

    if (length(maxcluster)>1){
    # build a null distribution by permuting signs/group labels
    for (n in 1:nresamples){
      tsum <- 0
      if (pracma::isempty(datay)){randsigns <- (round(runif(N))*2)-1}
      if (!pracma::isempty(datay) && is.matrix(datay)){randgroups <- permute(c(1:N,(N+1):(N+N2)))}
      for (elcounter in 1:length(maxcluster)){

        if (pracma::isempty(datay)){  # one-sample or paired test
          tempdataA <- datax[,maxcluster[elcounter]]*randsigns
          if (testtype==1){
            output <- t.test(tempdataA)
            tsum <- tsum + output$statistic}
          if (testtype==2){
            output <- tsqh.test(tempdataA)
            tsum <- tsum + output$tsq}
          if (testtype==3){
            output <- tsqc.test(tempdataA)
            tsum <- tsum + output$tsqc}
        }

        if (!pracma::isempty(datay) && is.matrix(datay)){  # independent samples test
          tempdata <- c(datax[,maxcluster[elcounter]],datay[,maxcluster[elcounter]])
          tempdataA <- tempdata[randgroups[1:N]]
          tempdataB <- tempdata[randgroups[(N+1):(N+N2)]]
          if (testtype==1){
            output <- t.test(tempdataA,tempdataB)
            tsum <- tsum + output$statistic}
          if (testtype==2){
            output <- tsqh.test(tempdataA,tempdataB,paired=FALSE)
            tsum <- tsum + output$tsq}
          if (testtype==3){
            output <- tsqc.test(tempdataA,tempdataB,paired=FALSE)
            tsum <- tsum + output$tsqc}
        }
      }
      nulldist[n] <- tsum
    }
    }

    # compare each cluster to the null distribution, retain the significant ones
    clusterps <- 0*(1:nclusts)
    if (length(nulldist)==nresamples){
    for (cc in 1:nclusts){
      i <- which(abs(nulldist)>abs(sumtvals[cc]))
      if (!isempty(i)){clusterps[cc] <- length(i)/nresamples}
      if (clusterps[cc]<clustthresh){
        sigclustercount <- sigclustercount + 1
        clusterpoints[[sigclustercount]] <- clusterlist[[cc]]
        clusterpout[sigclustercount] <- clusterps[cc]
        }
    }}
  }
clustout <- NULL
clustout$clusterpoints <- clusterpoints
clustout$nclusters <- sigclustercount
clustout$pvals <- clusterpout

return(clustout)
}

knitr::opts_chunk$set(echo = TRUE)
colpal <- c('#000000','#8783CF','#FE5000','#228B22','#6d008b','#8B8000')

@@

nbootstraps <- 1000    # number of bootstrap resamples for cluster correction
timeoffset <- 1000  # time offset for determining if suppression is increasing

@@

# Abstract

For decades, neural suppression in early visual cortex has been thought to be fixed. But recent work has challenged this assumption by showing that suppression can be \emph{reweighted} based on recent history; when pairs of stimuli are repeatedly presented together, suppression between them strengthens. Here we investigate the temporal dynamics of this process using a steady-state VEP paradigm that provides a time-resolved, direct index of suppression between pairs of stimuli flickering at different frequencies (5 and 7Hz). Our initial analysis of an existing EEG dataset (N=100) indicated that suppression increases substantially during the first 2-5 seconds of stimulus presentation (with some variation across stimulation frequency). We then collected new EEG data (N=100) replicating this finding for both monocular and dichoptic mask arrangements in a preregistered study designed to measure reweighting. A third experiment (N=20) used source localized MEG, and found that these effects are apparent in primary visual cortex (V1), consistent with results from neurophysiological work. Because long-standing theories propose inhibition/excitation differences in autism, we also compared reweighting between individuals with high vs low autistic traits, and with and without an autism diagnosis, across our 3 data sets (total N=220). We find no compelling differences in reweighting that are associated with autism. Our results support the normalization reweighting model, and indicate that for prolonged stimulation, increases in suppression occur on the order of 2-5 seconds after stimulus onset.

# Introduction

Suppressive interactions between neurons are ubiquitous in the nervous system, and normalization (or gain control) processes are considered a canonical neuronal computation [@Carandini2011]. Yet for decades the strength of suppression was treated as fixed, largely due to the observation that adapting to one stimulus does not decrease its suppressive potency [@Foley1997; @Freeman2002]. This orthodoxy was recently challenged by a series of innovative studies showing that normalization can be 'reweighted' by recent history [@Westrick2016; @Aschner2018; @Yiltiz2020]. Specifically, when pairs of stimuli are repeatedly presented together, their neural representations come to suppress each other more strongly. This suggests that, far from being fixed, normalization is a dynamic process that is continuously updated by the sensory environment. Here, our objectives were to determine if the timecourse of these changes can be measured non-invasively from the human brain, assess if they occur similarly in distinct suppressive pathways, and determine whether they differ across the population as a function of autistic traits.

Plastic changes within the visual system occur over multiple timescales [see @Webster2015 for a recent review]. In the retina, adaptation to complete darkness takes around 30 minutes [@Lamb2006], and has established physiological substrates (i.e. the regeneration rate of chemicals in different photoreceptor classes). Cortical forms of adaptation to cues such as stimulus contrast [@Blakemore1969], orientation [@Gibson1937] and motion [@Mather2008] can be observed within a few seconds, but also build up over durations on the order of several minutes [@Greenlee1991]. A recent form of binocular plasticity [@Lunghi2011], in which one eye is patched and subsequently contributes more to binocular vision, requires more than an hour. Other types of adaptation have been identified with changes occurring over longer time periods, such as over several hours [@Kwon2009] or days [@Haak2014]. Finally adaptation to specific combinations of orientation and colour [the @McCollough1965 effect] can persist for several weeks [@Jones1975], even with a brief induction period. Previous work on normalization reweighting has involved adapting sequences of around 40 -- 60s [@Aschner2018; @Yiltiz2020], though it is possible that reweighting might occur faster than this, consistent with other types of contrast adaptation.

Normalization itself is not a single process. Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation [@Heeger1992; @Foley1994], eye-of-origin [@Legge1979; @Sengpiel1994] and spatial position [@Petrov2005; @Cannon1991]. At present there is evidence of normalization reweighting between stimuli with orthogonal orientations [@Aschner2018], and adjacent spatial positions [@Yiltiz2020]. We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways. This is plausible, given that suppression within and between the eyes has different spatiotemporal tuning [@Meese2009], and dichoptic masking can be reduced by adapting to the mask [@Sengpiel2005; @Li2005; @Baker2007], unlike within-eye masking. Dynamic fluctuations in interocular suppression are a key feature of binocular rivalry [@Wilson2003; @Alais2010], for which autism-related differences have also been reported [@Robertson2013].

Atypical sensory experience is widely reported by individuals on the autism spectrum [@MacLennan2022; @Simmons2009], but the causal mechanisms remain unclear. Typical issues include hypersensitivity to intense stimuli such as loud sounds, bright lights and strong odours or flavours. Yet fundamental measures of sensitivity such as visual acuity [@Tavassoli2011], contrast sensitivity [@Koh2010], and audiometric performance [@Rosenhall1999] are not consistently different from neurotypical controls. Theoretical accounts of sensory differences in autism have long proposed that the balance of inhibition and excitation may be disrupted [@Rubenstein2003; @Rosenberg2015], and there are isolated results that appear consistent with this. For example neural measures of surround suppression are weaker in autism [@Schallmo2020], binocular rivalry is slower [@Robertson2013] and orientation discrimination performance is better [@Dickinson2014]. Conversely, other studies have failed to find group differences using paradigms that should be equally dependent on inhibition, such as orientation repulsion [@VandeCruys2018], repetition suppression [@Utzerath2018] and psychophysical measures of surround suppression [@Sandhu2020]. The evidence for atypical inhibition in autism is therefore currently inconclusive.

Our recent work has identified a potential autism-related difference in gain control using steady-state EEG methods. @Vilidaite2018 measured contrast response functions for flickering stimuli in a large group of neurotypical adults, who were split by their autism quotient [AQ, see @Baron-Cohen2001] score. The responses were comparable between high and low AQ groups at the flicker frequency of the stimulus. However at the second harmonic frequency (twice the flicker frequency), the high AQ group showed weaker responses than the low AQ group. This effect was replicated in an independent sample of 12 adults with an autism diagnosis (compared with neurotypical controls). Because second harmonic responses are caused by nonlinear interactions in the visual system (they are absent in a linear system), and divisive suppression is a nonlinearity, this potentially implicates differences in suppression in autism. Furthermore, there appears to be a developmental trajectory, as autistic children tested as part of the same study [@Vilidaite2018] showed weaker responses at both the first and second harmonic frequencies.

In this paper we perform a time-course analysis of a previously published EEG data set, and report two novel pre-registered experiments using EEG and MEG. Our data show that suppression increases substantially during the first 2-5 seconds following stimulus onset, for both monocular and dichoptic masks. Source localisation of MEG data indicate that the reweighting is present as early as primary visual cortex (V1). We also hypothesised that normalization reweighting might differ as a function of autistic traits. The relative novelty of the reweighting framework could explain why any differences have not previously been detected, and why the literature on inhibition in autism is relatively inconclusive. However we did not find convincing support for this hypothesis.

# Results

@@

earlydata <- meanindices*0
latedata <- meanindices*0
earlydata[,1:(dim(meanindices)[2]-(timeoffset/2))] <- meanindices[,1:(dim(meanindices)[2]-(timeoffset/2))]
earlydata[,1:(timeoffset/2)] <- 0
latedata[,(1+(timeoffset/2)):(dim(meanindices)[2])] <- meanindices[,1:(dim(meanindices)[2]-(timeoffset/2))]
latedata[1:dim(meanindices)[1],(dim(meanindices)[2])+(1-(timeoffset/2)):0] <- 0

if (nbootstraps>0){clustout <- clustercorrect(earlydata,latedata,testtype=4,paired=TRUE)}

@@

  if (nbootstraps>0){
    for (n in 1:clustout$nclusters){
      temp <- unlist(clustout$clusterpoints[n])
      lines(times[temp[c(1,length(temp))]]-((timeoffset/2)/1000),c(5.5, 5.5),lwd=3)
    }
  }

@@

  pdf('temp/figures/Fig1.pdf', bg="transparent", height = 6, width = 16)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

load('temp/Pilotsummary.RData')

targetcond <- NULL
maskcond <- NULL
subj <- NULL
targetamp <- NULL
maskamp <- NULL
i <- 0
for (s in 1:100){
  for (cond in 1:14){
    i <- i + 1
    subj[i] <- s
    targetcond[i] <- ((cond-1) %% 7) + 1
    maskcond[i] <- ceiling(cond/7)
    targetamp[i] <- abs(EEGallconds[s,1,cond])
    maskamp[i] <- abs(EEGallconds[s,2,cond])
  }
}
anovadata <- data.frame(subj,targetcond,maskcond,targetamp,maskamp)
anovadata$subj <- as.factor(anovadata$subj)
anovadata$targetcond <- as.factor(anovadata$targetcond)
anovadata$maskcond <- as.factor(anovadata$maskcond)

maskinganova1 <- ezANOVA(anovadata, dv=targetamp, wid=subj, within=c(targetcond,maskcond))

anovadataM <- subset(anovadata,maskcond==2)
maskinganova2 <- ezANOVA(anovadataM, dv=maskamp, wid=subj, within=targetcond)
gg <- maskinganova2$`Sphericity Corrections`$GGe

@@

We began by reanalysing data from a steady-state visually evoked potential (SSVEP) experiment reported by @Vilidaite2018. Participants viewed arrays of flickering gratings of varying contrasts. In some conditions a single grating orientation was present flickering at 7Hz, whereas in other conditions a high contrast 'mask' was added at right angles to the target gratings, and flickering at 5Hz. The left panel of Figure \@ref(fig:Pilotdata)a shows contrast response functions with and without the mask - the presence of the mask reduces the 7Hz response to the target (blue squares are below the black circles; significant main effect of mask contrast, F(1,99) = `r round(maskinganova1$ANOVA$F[2],digits=2)`, \emph{p} < 0.001). Similarly, the right panel of Figure \@ref(fig:Pilotdata)a shows that the 5Hz response to the mask was itself suppressed by the presence of high contrast targets (main effect of target contrast on the mask response, F(`r round(maskinganova2$ANOVA$DFn*gg,digits=2)`,`r round(maskinganova2$ANOVA$DFd*gg,digits=2)`) = `r round(maskinganova2$ANOVA$F,digits=2)`, \emph{p} < 0.001; note that the data from the masking conditions were not reported by @Vilidaite2018). At both frequencies, responses were localised to the occipital pole (see insets).

```{r Pilotdata, fig.cap="Summary of pilot analysis of data from Vilidiate et al. (2018). Panel (a) shows contrast response functions at the target frequency (7Hz, left) and the mask frequency (5Hz, right). Insets show the distribution of activity across the scalp, with points marking the electrodes over which signals were averaged (Oz, POz, O1 and O2). Panel (b) shows Fourier spectra for the single component stimuli and their combination (plaid). Note the strong second harmonic components at 14Hz and 10Hz. Panel (c) shows timecourses of frequency-locked responses to a single stimulus (black) and the plaid stimulus (blue), compared to baseline (grey). Panel (d) shows the timecourse of suppression at each frequency (7Hz, 5Hz, 14Hz, 10Hz) and their average (black curve). Points around y = 1.8 indicate a significantly increasing ratio. Error bars in panel (a) and shaded regions in panels (c,d) indicate ±1SE, and grey rectangles indicate the timing of stimulus presentation. The larger symbols in panel (a) indicate the conditions used for subsequent analyses.", fig.align="center", echo=FALSE}

knitr::include_graphics('temp/figures/Fig1.pdf')

@@

We then performed a timecourse analysis, in which we analysed each 11-second trial using a sliding 1-second time window. The top panel of Figure \@ref(fig:Pilotdata)c shows the response at the target frequency (7Hz) to a single stimulus of 32% contrast (black), and the response at 7Hz when the 32% contrast mask is added (blue). Analogous responses are shown at three other frequencies - the mask frequency (5Hz), and the second harmonics of both target and mask frequencies (14Hz, 10Hz), at which strong responses were also found (see spectra in Figure \@ref(fig:Pilotdata)b). The reduction in signal strength when the mask component is added illustrates the masking effect.

Taking the ratio of the two timecourses to calculate a masking index reveals that at 7Hz masking increases steeply during the first two seconds of stimulus presentation, and then plateaus for several seconds (blue trace in Figure \@ref(fig:Pilotdata)d). A similar pattern is observed at 5Hz (red trace in Figure \@ref(fig:Pilotdata)d), as well as at the second harmonics, with some variability in the timecourse across frequencies; for example at 5Hz suppression peaks at around 4 seconds. The black trace shows the average masking ratio across all four frequencies, which rises steeply for just over two seconds and then stays approximately constant until stimulus offset. We conducted cluster-corrected t-tests between ratios separated by `r timeoffset`ms, testing for an increase in suppression ratio across time (i.e. a one-sided test). Points at y = 1.8 in Figure \@ref(fig:Pilotdata)d indicate time points where the ratio is significantly increasing, and occur mostly during the first 2 seconds of stimulus presentation.

Our initial reanalysis was promising, however the data were noisy despite the large sample size, because each participant contributed only 8 trials (88 seconds) to each condition. We therefore preregistered two new experiments (see https://osf.io/4qudc) to investigate these effects in greater detail. These had a similar overall design to the @Vilidaite2018 study, with some small changes intended to optimise the study (see Methods). The key differences were that we used shorter trials (because there were few changes in the latter part of the trials shown in Figure \@ref(fig:Pilotdata)d), and also focussed all trials into a smaller number of conditions, such that each participant contributed 48 repetitions (288 seconds of data) to each of 4 conditions.

@@

allAQ[s] <- scoreAQ(qdatasubj[1,5:54]) 
allSPQ[s] <- scoreSPQ(qdatasubj[1,55:146])
}

save(file='temp/EEGsummary.RData',list=c('allheads5','allheads7','allspec','allwaves','EEGtimecourse5Hz','EEGtimecourse7Hz','electrodes','allAQ','allSPQ'))
 }

load('temp/EEGsummary.RData')

@@

earlydataM <- monratios*0
latedataM <- monratios*0
earlydataM[,1:(dim(monratios)[2]-(timeoffset/2))] <- monratios[,1:(dim(monratios)[2]-(timeoffset/2))]
earlydataM[,1:(timeoffset/2)] <- 0
latedataM[,(1+(timeoffset/2)):(dim(monratios)[2])] <- monratios[,1:(dim(monratios)[2]-(timeoffset/2))]
latedataM[1:dim(monratios)[1],(dim(monratios)[2])+(1-(timeoffset/2)):0] <- 0

if (nbootstraps>0){clustoutM5 <- clustercorrect(earlydataM,latedataM,testtype=4,paired=TRUE)}
